Regression models for Cylindrical data in Psychology

### 2019-02-02 Manuscript/Manuscript.Rmd
Manuscript Update

header-includes:
  - \DeclareRobustCommand{\VANDER}[3]{#2} 
  - \usepackage{multirow}
  - \usepackage{rotating}
  - \usepackage{color}
  - \usepackage{subcaption}
   #% set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

>>>>

header-includes:
  - \DeclareRobustCommand{\VANDER}[3]{#2} 
  - \DeclareRobustCommand{\VAN}[3]{#2}
  - \DeclareRobustCommand{\DEN}[3]{#2}
  - \usepackage{multirow}
  - \usepackage{rotating}
  - \usepackage{color}
  - \usepackage{subcaption}
   #% set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

### 2019-02-02 Manuscript/Manuscript.tex
Manuscript Update

\linenumbers
\DeclareRobustCommand{\VANDER}[3]{#2}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{color}
\usepackage{subcaption}

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006), the circumplex of affect (Russell, 1980), the
circumplex of human emotion (Plutchik, 1997) or the model of human
values (Schwartz, 1992). \newline \indent Also, many of the more recent
types of data that are studied in psychology are cylindrical. For
example, research on human navigation uses data where distance (i.e.,
linear component) and direction (i.e., the linear component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data
also consist of both the direction (i.e., the circular variable) and the
duration (i.e., the linear variable)(e.g., Rayner (2009)). Apart from
the social sciences data with a circular and linear outcome more
commonly occur in meteorology (García-Portugués, Crujeiras, \&
González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras,
González-Manteiga, \& Pereira, 2014) or marine research (Lagona, Picone,
Maruotti, \& Cosoli, 2015) \newline \indent Up until now researchers
studying cylindrical data had to rely on linear statistical methods to
analyze their research results. However, lately more and more of these
researchers acknowledge that linear methods are not sufficient and call
for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy,
\& Hilsenroth, 2009) that take into account both the circular and the
linear component of these data. \newline \indent The aim in the present
paper is twofold. Firstly, we we intend to fill the above mentioned gap
in the literature by showing the use of cylindrical models can benefit
the analysis of circumplex data and cylindrical data in psychology in
general. More specifically we will show these benefits for interpersonal
teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components
from a set of covariates in a regression model. Our second aim therefore
is to adapt several existing cylindrical models in such a way that they
include a regression of both the linear and circular component of a
cylindrical variable onto a set of covariates. From now on we will
therefore refer to the components of the cylindrical variable as outcome
components. These adapted circular models are then used to analyse the
teacher data.

@@

@@

\leavevmode\hypertarget{ref-pennings2018interpersonal}{}%
Pennings, H. J. M., Brekelmans, M., Sadler, P., Claessens, L. C.,
\VANDER{Want}{Van der}{van der} Want, A. C., \& Tartwijk, J. van.
(2018). Interpersonal adaptation in teacher-student interaction.
\emph{Learning and Instruction}, \emph{55}, 41--57.
doi:\href{https://doi.org/10.1016/j.learninstruc.2017.09.005}{10.1016/j.learninstruc.2017.09.005}

@@

\leavevmode\hypertarget{ref-want2018selfefficacy}{}%
\VANDER{Want}{Van der}{van der} Want, A. C., Den Brok, P., Beijaard, D.,
Brekelmans, M., Claessens, L. C., \& Pennings, H. J. M. (2018). The
relation between teachers' interpersonal role identity and their
self-efficacy, burnout and work engagement. \emph{Professional
Development in Education}, Advance Online Publication.
doi:\href{https://doi.org/10.1080/19415257.2018.1511453}{10.1080/19415257.2018.1511453}

@@

\leavevmode\hypertarget{ref-wubbels2006interpersonal}{}%
Wubbels, T., Brekelmans, M., Brok, P. den, \& Tartwijk, J. van. (2006).
Handbook of classroom management: Research practice and contemporary
issues. In C. Evertson \& C. S. Weinstein (Eds.), \emph{Handbook of
classroom management: Research, practice, and contemporary issues} (pp.
1161--1191). Malwah, NJ: Lawrence Erlbaum Associates.

>>>>

\linenumbers
\DeclareRobustCommand{\VANDER}[3]{#2}
\DeclareRobustCommand{\VAN}[3]{#2}
\DeclareRobustCommand{\DEN}[3]{#2}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{color}
\usepackage{subcaption}

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
\DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006),
the circumplex of affect (Russell, 1980), the circumplex of human
emotion (Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline \indent Also, many of the more recent types of data that are
studied in psychology are cylindrical. For example, research on human
navigation uses data where distance (i.e., linear component) and
direction (i.e., the linear component) are of interest (Chrastil \&
Warren, 2017) or in eye-tracking, the saccade data also consist of both
the direction (i.e., the circular variable) and the duration (i.e., the
linear variable)(e.g., Rayner (2009)). Apart from the social sciences
data with a circular and linear outcome more commonly occur in
meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013),
ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \&
Pereira, 2014) or marine research (Lagona, Picone, Maruotti, \& Cosoli,
2015) \newline \indent Up until now researchers studying cylindrical
data had to rely on linear statistical methods to analyze their research
results. However, lately more and more of these researchers acknowledge
that linear methods are not sufficient and call for new methods
(Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth,
2009) that take into account both the circular and the linear component
of these data. \newline \indent The aim in the present paper is twofold.
Firstly, we we intend to fill the above mentioned gap in the literature
by showing the use of cylindrical models can benefit the analysis of
circumplex data and cylindrical data in psychology in general. More
specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the
dependence between the linear and circular component of a cylindrical
variable we would also like to predict the two components from a set of
covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a
regression of both the linear and circular component of a cylindrical
variable onto a set of covariates. From now on we will therefore refer
to the components of the cylindrical variable as outcome components.
These adapted circular models are then used to analyse the teacher data.

@@

\DeclareRobustCommand{\VANDER}[3]{#3}
\DeclareRobustCommand{\VAN}[3]{#3}
\DeclareRobustCommand{\DEN}[3]{#3}

@@

\leavevmode\hypertarget{ref-pennings2018interpersonal}{}%
Pennings, H. J. M., Brekelmans, M., Sadler, P., Claessens, L. C.,
\VANDER{Want}{Van der}{van der} Want, A. C., \& \VAN{Tartwijk}{Van}{van}
Tartwijk, J. (2018). Interpersonal adaptation in teacher-student
interaction. \emph{Learning and Instruction}, \emph{55}, 41--57.
doi:\href{https://doi.org/10.1016/j.learninstruc.2017.09.005}{10.1016/j.learninstruc.2017.09.005}

@@

\leavevmode\hypertarget{ref-want2018selfefficacy}{}%
\VANDER{Want}{Van der}{van der} Want, A. C., \DEN{Brok}{Den}{den} Brok,
P., Beijaard, D., Brekelmans, M., Claessens, L. C., \& Pennings, H. J.
M. (2018). The relation between teachers' interpersonal role identity
and their self-efficacy, burnout and work engagement. \emph{Professional
Development in Education}, Advance Online Publication.
doi:\href{https://doi.org/10.1080/19415257.2018.1511453}{10.1080/19415257.2018.1511453}

@@

\leavevmode\hypertarget{ref-wubbels2006interpersonal}{}%
Wubbels, T., Brekelmans, M., \DEN{Brok}{Den}{den} Brok, P., \&
\VAN{Tartwijk}{Van}{van} Tartwijk, J. (2006). Handbook of classroom
management: Research practice and contemporary issues. In C. Evertson \&
C. S. Weinstein (Eds.), \emph{Handbook of classroom management:
Research, practice, and contemporary issues} (pp. 1161--1191). Malwah,
NJ: Lawrence Erlbaum Associates.
